---
title: 'Manipulating and Plotting Data in R: A Quick Start Guide'
author: "Joshua Keller"
date: "8/15/2020"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 5)
knitr::opts_chunk$set(fig.height = 3.5)
```


# About

This document provides a reference for getting started witht data manipulation and visualization tasks in R. It is not even close to being exhaustive and represents only a very small fraction of what is possible in R. 

A recommend resource with more information is R for Data Science (<https://r4ds.had.co.nz>) by Wickham and Grolemund.


# Loading Packages

R packages provide sets of functions designed for completing specific tasks. To load a package, use the `library()` command:

```{r message=FALSE}
library(tidyverse)
library(palmerpenguins)
```

The package we will use the most is the `tidyverse` package, which is a collection of many different functions for reading, manipulating, and visualizing data. More information on the tidyverse is available at <https://www.tidyverse.org>.

# Reading Data

There are three primary ways to load data into R, depending on its form.

## Comma-separated value (.csv) files

For a `.csv` file, use the `read_csv()` command. 

* The first argument is the (relative) path to the file.
* Assign the output to your desired variable name

```{r}
income <- read_csv("../data/median_income.csv")
```

## Data provided in packages

Some R packages, such as the `palmerpenguins` package, come with data included. In that case, you just refer to the data frame by name:

```{r}
penguins
```

## Data stored in an `RData` file

R objects can be saved as `.RData` files. To load them into your workspace, use the `load()` command:

```{r include=F, eval=F}
testdata <- 1:10
save(testdata, file="../data/testdata.RData")
```


```{r}
load("../data/testdata.RData", verbose=TRUE)
```


# Viewing Data Objects

The simplest way to view part of a dataset is to just type its name in the console:

```{r}
income
```

For data frames and tibbles, this will print the first several lines. For other objects, it will print a summary of the object information.


You can also view objects by clicking on their name in the "Environment" pane of the RStudio IDE.

\newpage 
# Plotting Data

The primary function for plotting data is `ggplot()`, which comes in the `ggplot2` package and is also part of `tidyverse`.

`ggplot()` works by "adding" together layers.

* The first layer is always `ggplot()`.
* The next layer can be a `geom_*` function, which controls the type of plot made.
    * `geom_point()` is used to make a scatterplot
    * `geom_histogram()` is used to make a histogram
    * `geom_abline()` is used to make a line, using a provided slope and intercept
    * many other functions exist

Within each `geom_*` function, we need to provide:

* The `aes()` command that specifies the variables used as the `x` and/or `y` values
    * The variable names do not need quotes.
* A `data=` argument that specifies the data frame or tibble containing the necessary variables.

## Scatterplot

Example code for a scatter plot is:
```{r warning=FALSE}
ggplot() + 
    geom_point(aes(x=flipper_length_mm,
                   y=body_mass_g),
               data=penguins)
```

We can make the background white instead of gray by adding a  `theme_bw()` layer:

```{r warning=FALSE}
ggplot() + 
    theme_bw() + 
    geom_point(aes(x=flipper_length_mm,
                   y=body_mass_g),
               data=penguins)
```


### Adding color

If we want to use a categorical variable to change the color of the plot, we can set `col=` *inside* the `aes()` command:

```{r warning=FALSE}
g_penguin <- ggplot() + 
    theme_bw() + 
    geom_point(aes(x=flipper_length_mm,
                   y=body_mass_g,
                   col=species),
               data=penguins)
g_penguin
```

Note also here that we have assigned the plot to an object (`g_penguin`). We can then print the plot or add to it by calling that object name.


Instead, if we want to only change the color of *all* points, set `col=` **outside** of `aes()`:

```{r warning=FALSE}
ggplot() + 
    theme_bw() + 
    geom_point(aes(x=flipper_length_mm,
                   y=body_mass_g),
                   col="blue",
               data=penguins)
```



### Adding axis labels and title

* To add an axis label, use `+ xlab("Label here")` or `ylab("Y label here")`.
* To add a title, use `ggtitle("Title here")`

```{r warning=FALSE}
g_penguin + 
    xlab("Flipper Length (mm)") +
    ylab("Body Mass (g)") +
    ggtitle("Palmer Penguin Data")
```


### Adding a line

* To add a line with known slope and intercept, use the `geom_abline()` command.
    * It requires `slope=` and `intercept=` inside the `aes()` command.

```{r warning=FALSE}
g_penguin + 
    xlab("Flipper Length (mm)") +
    ylab("Body Mass (g)") +
    theme_bw() + 
    ggtitle("Palmer Penguin Data") +
    geom_abline(aes(slope=10,
                    intercept=2000))
```

## Histogram

* Histograms can be made using the `geom_histogram()` command.
    * This takes only `x=` (but not `y=`) inside its `aes()`.
    * The number of bins can be specified using `bins=`.

```{r warning=FALSE}
ggplot() + 
    geom_histogram(aes(x=flipper_length_mm),
                   bins=20,
                   data=penguins) +
    xlab("Flipper Length (mm)") + 
    ggtitle("Penguin Flipper Length")
```

### Adding Color

Adding color to histograms should be done with care, since it can be difficult to read.

To change the color of the histogram as whole, set `fill=` outside of `aes()`. (For a histogram, `col=` referes to the outline of the boxes, while `fill=` refers to the color within the bars.)

```{r warning=FALSE}
ggplot() + 
    theme_bw() + 
    geom_histogram(aes(x=flipper_length_mm),
                   bins=20,
                   fill="red",
                   data=penguins) +
    xlab("Flipper Length (mm)") + 
    ggtitle("Penguin Flipper Length")
```


To create overlapping histograms each of a different color, do all of the following:

* set `fill=` inside of `aes()` 
* set `position="identity"` in the `geom_histogram()` command.
* set `alpha=0.5` (or other number between 0.2 and 0.5) in the `geom_histogram()` command. This makes the colors partially transparent, which allows you to see the overlap.

```{r warning=FALSE}
ggplot() + 
    theme_bw() + 
    geom_histogram(aes(x=flipper_length_mm,
                       fill=species),
                   bins=20,
                   alpha=0.5,
                   position="identity",
                   data=penguins) +
    xlab("Flipper Length (mm)") + 
    ggtitle("Penguin Flipper Length")
```


## Boxplot

* Boxplots can be made using the `geom_boxplot()` command.
    * To make a horizontal boxplot, use `x=` inside `aes()`.
    * To make a vertical boxplot use `y=` inside `aes()`.

```{r warning=FALSE, fig.width=3, fig.height=5}
ggplot() + 
    theme_bw() + 
    geom_boxplot(aes(y=flipper_length_mm),
                   data=penguins) +
    ylab("Flipper Length (mm)") + 
    ggtitle("Penguin Flipper Length")
```




# Manipulating Data

Many different approaches to calculating data summaries in R. We will primarily use functions from `tidyr` and `dplyr` packages, both of which are in the `tidyverse`.

* Functions operate on data frames and tibbles
* Link functions together using the "pipe": %>%
* Key functions are: `group_by`, `mutate`, and `summarize`


## Summarizing Data in R

\texttt{summarize()} applies a function to all observations

```{r echo=TRUE}
summarize(penguins, mean(flipper_length_mm, na.rm=T))
```

When using `%>%`, the first argument is automatically passed forward
```{r}
penguins %>%
  summarize(mean_flipper_length=mean(flipper_length_mm, na.rm=T))
```
\texttt{n()} counts the number of observations

```{r}
penguins %>%
  summarize(n=n())
```


\texttt{summarize()} can take multiple functions at once

```{r echo=TRUE}
penguins %>%
  summarize(n=n(),
            mean_flipper_length=mean(flipper_length_mm, na.rm=T),
            `Mean Body Mass (g)` = mean(body_mass_g, na.rm=T))
```

Set `group_by` to have `summarize()` calculate values within each group.

```{r echo=TRUE}
penguins %>%
    group_by(species) %>%
  summarize(n=n(),
            mean_flipper_length=mean(flipper_length_mm, na.rm=T),
            `Mean Body Mass (g)` = mean(body_mass_g, na.rm=T))
```

* `group_by` can take multiple variables

```{r echo=TRUE}
penguins %>%
    group_by(species, sex) %>%
  summarize(mean_flipper_length=mean(flipper_length_mm, na.rm=T),
            `Mean Body Mass (g)` = mean(body_mass_g, na.rm=T))
```

## Creating/Changing variables
 
\texttt{mutate()} creates/replaces variables


```{r echo=TRUE, size="footnotesize"}
names(penguins)
penguins <- penguins %>%
  mutate(flipper_length_in=flipper_length_mm/25.4)
names(penguins)
penguins %>%
    select(flipper_length_in, flipper_length_mm)
```



## Summarizing Data in R

\texttt{group\_by()} is often combined with \texttt{n()} 

```{r}
mtcars %>%
  group_by(cyl) %>%
  summarize(n=n()) %>%
  mutate(freq = n / sum(n))
```

## Summarizing Data in R

```{r}
mtcars %>%
  group_by(cyl) %>%
  summarize(n=n()) %>%
  mutate(freq = round(n / sum(n), 2))
```



## Tables

\texttt{table()} is robust and useful for (cross-)tabulating  categorical variables. It is not part of `tidyverse`, so doesn't work well as with `summarize` or `group_by`.

```{r}
t1 <- table(penguins$species, penguins$sex)
t1
addmargins(t1)
```

Similar behavior can be achieved in tidyverse, with a little more work:
```{r warning=FALSE}
penguins %>%
  group_by(species, sex) %>%
  summarize(n=n()) %>%
  spread(species, n)
```

